Preferences for the commodification of pensions in Chile: the rol of intergenerational social mobility

### 2025-11-22 paper/paper.tex
Up V2

\setstretch{1.5}
This document was last modified at 2025-11-21 16:32:59 and it was last
rendered at 2025-11-21 16:32:59.

@@

What is the legitimate scope of market inequality in the eyes of the
public? Since the early 1980s, many countries have experienced a
rollback of universal welfare programs and a shift toward the
privatisation and commodification of public goods and social services
(\citeproc{ref-gingrich_making_2011}{Gingrich, 2011};
\citeproc{ref-streeck_how_2016}{Streeck, 2016}). In Latin America, these
reforms extended market logic into domains of social reproduction long
governed by the state, shrinking public provision and expanding private
actors in core welfare sectors (\citeproc{ref-ferre_welfare_2023}{Ferre,
2023}). From a moral-economy perspective, this diffusion of market rules
has reconfigured the normative order: through policy feedback, welfare
institutions increasingly embed market-based criteria of ``fair''
allocation and shape how citizens understand deservingness and the
balance between state and market (\citeproc{ref-koos_moral_2019}{Koos \&
Sachweh, 2019};
\citeproc{ref-svallforsMoralEconomyClass2006a}{Svallfors, 2006}).
Against this backdrop, growing scholarship examines \emph{market justice
preferences}---the extent to which citizens regard market-based criteria
as a fair basis for access to essential services such as healthcare,
education and pensions
(\citeproc{ref-castillo_perceptions_2025}{Castillo, Laffert, et al.,
2025}; \citeproc{ref-lindh_public_2015}{Lindh, 2015};
\citeproc{ref-svallfors_political_2007}{Svallfors, 2007}). These
orientations matter because they help legitimise economic inequality by
portraying stratified outcomes as a fair result of individual
responsibility and limited state intervention.

This article examines public support for the commodification of pension
welfare---specifically, whether people consider it fair that access to
better pensions depends on income or contributions. Pension
privatisation lies at the core of Latin America's trajectory of
commodification (\citeproc{ref-huber_political_2000}{Huber \& Stephens,
2000}), making pension systems a strategic domain for analysing how
market principles become normatively embedded
(\citeproc{ref-mau_inequality_2015}{Mau, 2015}). As many countries
expanded mandatory or voluntary individual-capitalisation schemes---most
radically in Chile---pensions became organised around contribution-based
entitlements and market performance
(\citeproc{ref-arenas_sistemas_2019}{Arenas, 2019};
\citeproc{ref-verbic_political_2019}{Verbič \& Spruk, 2019}). This shift
reshapes not only how pension institutions allocate old-age risk,
heightening exposure to poverty in retirement and tying benefit levels
to lifetime labour-market trajectories
(\citeproc{ref-madero-cabib_private_2019}{Madero-Cabib et al., 2019};
\citeproc{ref-oecd_pensions_2023}{OECD, 2023}), but also reshape the
justice principles citizens associate with old-age security: people have
to judge whether higher benefits should follow market-derived criteria
or solidaristic principles
(\citeproc{ref-arrizabalomontoro_debate_2019}{Arrizabalo Montoro et al.,
2019}; \citeproc{ref-borzutzky_pension_2012}{Borzutzky, 2012}). Although
we know something about what people consider a ``just'' monetary pension
(\citeproc{ref-castillo_deserving_2019}{Castillo, Olivos, et al., 2019})
and about support for public versus private provision
(\citeproc{ref-busemeyer_welfare_2020}{Busemeyer \& Iversen, 2020};
\citeproc{ref-jaime-castillo_public_2013}{Jaime-Castillo, 2013}), we
still lack systematic evidence on how citizens assess the fairness of
market-dependent access to pensions.

Research shows that support for market-based access to welfare services
is strongly stratified: individuals with higher income, education or
class position are more likely to endorse market-based criteria than
disadvantaged groups (\citeproc{ref-busemeyer_skills_2014}{Busemeyer,
2014}; \citeproc{ref-immergut_it_2020}{Immergut \& Schneider, 2020};
\citeproc{ref-lindh_public_2015}{Lindh, 2015};
\citeproc{ref-svallfors_political_2007}{Svallfors, 2007}). Evidence for
pensions points in the same direction: higher income and greater
financial gains in funded schemes correlate with stronger support for
market-oriented pension policies
(\citeproc{ref-kerner_pension_2020}{Kerner, 2020}). Yet we know very
little about how movement within the class structure shapes preferences
for pension commodification. This gap is especially salient in Latin
America, where mobility has unfolded under high inequality and deeply
privatised welfare regimes
(\citeproc{ref-lopez-roldan_comparative_2021}{López-Roldán \& Fachelli,
2021}). Social origins and destinations anchor material interests and
justice principles
(\citeproc{ref-alesina_intergenerational_2018}{Alesina et al., 2018};
\citeproc{ref-gugushvili_subjective_2017}{Gugushvili, 2017};
\citeproc{ref-langsaether_explaining_2022}{Langsæther et al., 2022}),
while movement between these positions exposes individuals to new risks,
resources and normative environments that reshape fairness judgements
(\citeproc{ref-gugushvili_trends_2014}{Gugushvili, 2014};
\citeproc{ref-helgason_longterm_2023}{Helgason \& Rehm, 2023}). As Mau
(\citeproc{ref-mau_inequality_2015}{2015}) argues, social mobility can
foster market-justice orientations, helping explain how inequalities in
access to welfare services become normatively sustained.

The literature identifies several mechanisms through which mobility may
shape distributive-justice preferences---material self-interest
(\citeproc{ref-benabou_social_2001}{Benabou \& Ok, 2001}), socialisation
and acculturation
(\citeproc{ref-jaime-castillo_social_2019}{Jaime-Castillo \&
Marqués-Perales, 2019}), learning through long-term exposure
(\citeproc{ref-helgason_longterm_2023}{Helgason \& Rehm, 2023}), and
cognitive attributions such as self-serving bias or privilege awareness
(\citeproc{ref-gugushvili_intergenerational_2016c}{Gugushvili, 2016a};
\citeproc{ref-molina_its_2019}{Molina et al., 2019}). While distinct,
these accounts converge in highlighting two broad channels relevant for
market-based welfare: shifts in material stakes and shifts in normative
frames (\citeproc{ref-mau_inequality_2015}{Mau, 2015}). In this paper, I
do not treat these mechanisms as mediators but use them to guide
hypotheses about heterogeneity: if mobility alters preferences through
material interests and normative updating, its effects should be most
potent among those who interpret social outcomes through merit-based
evaluations of effort
(\citeproc{ref-castillo_perceptions_2025}{Castillo, Laffert, et al.,
2025}; \citeproc{ref-mijs_belief_2022}{Mijs et al., 2022}).

Against this background, this article pursues two aims: first, to assess
whether intergenerational mobility causally affects support for the
commodification of pensions; and second, to examine whether this effect
varies by meritocratic beliefs and over time. I argue that mobility
works through two main channels. Upward mobility increases earnings,
expected returns in a funded system and perceived economic security,
making contribution-based access more appealing. At the same time,
integration into higher-class networks exposes people to more
individualistic norms and effort-based understandings of fairness.
Downward mobility, in turn, heightens economic insecurity, dependence on
collective protection and attachment to solidaristic principles,
lowering support for market allocation. Meritocratic beliefs are
conceptualised not as mediators but as moderators. Among those who see
mobility as a fair reward for effort, both material and normative
effects should be amplified, producing higher support for pension
commodification. The paper contributes to research on market justice by
focusing on a specific welfare domain and by providing causal evidence
on how mobility shapes distributive preferences.

This study focuses on Chile, a particularly instructive case for
examining preferences for pension commodification. Despite sustained
economic growth, Chile combines high and persistent inequality with
short-range upward mobility and substantial barriers to higher-class
positions (\citeproc{ref-flores_top_2020}{Flores et al., 2020};
\citeproc{ref-salgado_uplifting_2025}{Salgado et al., 2025};
\citeproc{ref-torche_intergenerational_2014}{Torche, 2014}). It was the
first country worldwide to fully replace a public pay-as-you-go system
with a mandatory, privately administered (through pension fund
administrators, AFPs) defined-contribution scheme, while the state has
progressively expanded a segmented solidarity pillar to compensate for
insufficient benefits (\citeproc{ref-boccardo_30_2020}{Boccardo, 2020};
\citeproc{ref-madariaga_three_2020}{Madariaga, 2020};
\citeproc{ref-solimano_rise_2021}{Solimano, 2021}). In parallel---and
despite waves of protest from 2006 to 2019 targeting the AFP system
(\citeproc{ref-somma_no_2021}{Somma et al., 2021})---Chilean
subjectivities have been increasingly shaped by neoliberal discourses
and market logics, influencing attitudes toward pension distribution and
inequality (\citeproc{ref-araujo_desafios_2012}{Araujo \& Martuccelli,
2012}; \citeproc{ref-galvez_con_2023}{Gálvez, 2023}). The combination of
profound inequality, market-driven old-age security and contested
legitimacy makes Chile a critical case for analysing how mobility and
meritocratic beliefs structure support for pension commodification.

To address these questions, the study draws on large-scale,
representative survey data for the urban Chilean population collected in
2016, 2018 and 2023. It adopts a causal framework of mobility effects
inspired by Breen \& Ermisch (\citeproc{ref-breen_effects_2024}{2024}).
The following section outlines the theoretical framework linking
preferences for market-based welfare, social mobility and expected
heterogeneity by meritocratic beliefs and time, and derives hypotheses.
A subsequent section presents the causal identification strategy, the
data, the variables, and the analytical approach. The empirical sections
report the findings and interpret their implications. The article
concludes by discussing what these results reveal about the politics of
market-based pension welfare and social mobility.

@@

\subsection{Market justice
preferences}\label{market-justice-preferences}

@@

To test heterogeneity in mobility effects, I examine two moderators:
meritocratic perceptions and time. Meritocracy is captured with two
items (\citeproc{ref-young_rise_1958}{Young, 1958}), one referring to
effort (``In Chile, people are rewarded for their efforts'') and one to
talent (``In Chile, people are rewarded for their intelligence and
skills''), each answered on a five-point Likert scale (1 = ``strongly
disagree'' to 5 = ``strongly agree''). I average the two items into a
single index and dichotomize it into low meritocracy (≤ 3) and high
meritocracy (≥ 4). Time is measured by survey wave (2016, 2018, 2023)
and entered as a categorical (dummy) variable. Both variables are used
as moderators by interacting them with the mobility treatment to assess
conditional average treatment effects---i.e., whether the effect of
mobility varies across levels of meritocratic perceptions or across
survey waves.

@@

The specification is doubly robust: entropy-balancing weights are
combined with covariate adjustment so consistency holds if either the
weighting model or the outcome model is correctly specified, while also
improving efficiency
(\citeproc{ref-wooldridge_introductory_2009}{Wooldridge, 2009}). These
results are virtually identical to baseline models without covariates
(see \hyperref[anexo]{Supplementary Material} for complete models), with
no substantive changes in the mobility coefficients. Heterogeneous
effects by meritocratic perceptions and survey wave are examined through
interaction models described in the previous section.

@@

\subsection{Effect heterogeneity}\label{effect-heterogeneity-1}

To probe potential mechanisms underlying the mobility effects, I
estimate origin-specific ATT models that interact the mobility treatment
with high meritocratic perceptions (ref. = low), using stabilized IPW
weights, covariate adjustment, and wave fixed effects. I also estimate
an analogous specification interacting mobility with survey waves (ref.
= 2016) to assess temporal heterogeneity in the treatment effect (see
\hyperref[anexo]{Supplementary Material} for complete models).

@@

Figure~\ref{fig-interact} displays the origin-specific ATT estimates
from the meritocracy interaction models as marginal effects of the
mobility treatment at low and high meritocratic beliefs. For low-merit
respondents, mobility effects are uniformly small and imprecise across
trajectories: Low→Middle (\(\beta\) = −0.02, 95\% CI {[}−0.09, 0.04{]}),
Low→High (\(\beta\) = −0.03, {[}−0.10, 0.05{]}), Middle→Low (\(\beta\) =
−0.00, {[}−0.07, 0.06{]}), and a modest, non-significant positive
estimate for Middle→High (\(\beta\) = 0.07, {[}−0.01, 0.14{]}), while
downward moves from high origin show borderline negative effects
(High→Middle: \(\beta\) = −0.07, {[}−0.15, 0.00{]}; High→Low: \(\beta\)
= −0.08, {[}−0.15, 0.00{]}). Among high-merit respondents, the pattern
is directionally consistent with meritocratic moderation but
statistically robust only for one trajectory: the ATT for Middle→High at
high merit is positive and significant (\(\beta\) = 0.22, {[}0.05,
0.38{]}), whereas Low→Middle (\(\beta\) = 0.02, {[}−0.13, 0.16{]}),
Low→High (\(\beta\) = 0.02, {[}−0.15, 0.20{]}), High→Middle (\(\beta\) =
−0.10, {[}−0.28, 0.09{]}), and High→Low (\(\beta\) = −0.22, {[}−0.44,
0.00{]}) remain statistically indistinguishable from zero. Overall,
these results provide only limited and trajectory-specific evidence that
meritocratic beliefs moderate the causal effect of intergenerational
occupational mobility on preferences for pension commodification, with a
clearer pro-commodification sign only for upward mobility from middle to
high status among high-merit respondents.

\includegraphics[width=0.9\linewidth,height=\textheight,keepaspectratio]{paper_files/figure-pdf/fig-interact-wave-1.pdf}

\caption{\label{fig-interact-wave}Effects of intergenerational
occupational mobility on preferences for pension commodification, by
survey wave. Coefficient plot of origin-specific ATT estimates comparing
mobility versus immobility, estimated from interaction models (T ×
Wave). Points show the marginal effects (simple slopes) of the mobility
treatment in 2016, 2018, and 2023; bars are 95\% confidence intervals.
Estimates from pooled WLS with doubly robust adjustment, wave fixed
effects, and standard errors clustered by individual.}

I find no consistent evidence of heterogeneous causal effects across
survey waves. Figure~\ref{fig-interact-wave} displays the marginal
effects of mobility trajectories on market-justice preferences for
pensions in 2016, 2018, and 2023. Across trajectories, the estimated
ATTs remain small and their confidence intervals largely overlap over
time. Upward mobility from middle origin shows a positive and
statistically significant effect in 2016 (Middle→High, \(\beta\) = 0.13,
95\% CI {[}0.05, 0.22{]}) and downward mobility from high origin
displays negative effects at baseline (High→Middle and High→Low, both
\(\beta\) ≈ −0.19, CIs excluding zero), but these patterns attenuate and
become statistically indistinguishable from zero by 2023. Taken
together, the results suggest that any initial pro-commodification
response to upward mobility and de-commodifying response to downward
mobility are not stable over time, and the causal effect of mobility on
preferences for pension commodification is effectively time-invariant
over the 2016--2023 period.

@@

\caption{\label{tbl-complete3}Marginal effects of mobility by
meritocratic perception (Low vs High Meritocracy).}

@@

@@

\caption{\label{tbl-complete4}Marginal effects of mobility by survey
wave (2016, 2018, and 2023).}

@@

\centering
\begin{tabular}{l|c|c|c|c|c}
\hline
Trajectory & Wave & Estimate & Std. Error & CI Low & CI High\\
\hline
Low-Middle & 2016 & -0.048 & 0.055 & -0.156 & 0.060\\
\hline
Low-Middle & 2018 & -0.022 & 0.048 & -0.116 & 0.072\\
\hline
Low-Middle & 2023 & 0.022 & 0.058 & -0.092 & 0.135\\
\hline
Low-High & 2016 & -0.025 & 0.062 & -0.146 & 0.096\\
\hline
Low-High & 2018 & -0.088 & 0.050 & -0.186 & 0.011\\
\hline
Low-High & 2023 & 0.072 & 0.066 & -0.058 & 0.202\\
\hline
Middle-High & 2016 & 0.132 & 0.044 & 0.047 & 0.218\\
\hline
Middle-High & 2018 & 0.091 & 0.051 & -0.008 & 0.190\\
\hline
Middle-High & 2023 & 0.045 & 0.061 & -0.074 & 0.164\\
\hline
Middle-Low & 2016 & 0.061 & 0.044 & -0.025 & 0.147\\
\hline
Middle-Low & 2018 & -0.027 & 0.046 & -0.118 & 0.064\\
\hline
Middle-Low & 2023 & -0.010 & 0.057 & -0.122 & 0.103\\
\hline
High-Middle & 2016 & -0.187 & 0.070 & -0.324 & -0.050\\
\hline
High-Middle & 2018 & -0.078 & 0.050 & -0.175 & 0.019\\
\hline
High-Middle & 2023 & 0.006 & 0.063 & -0.118 & 0.130\\
\hline
High-Low & 2016 & -0.186 & 0.068 & -0.319 & -0.053\\
\hline
High-Low & 2018 & -0.125 & 0.053 & -0.228 & -0.022\\
\hline
High-Low & 2023 & -0.009 & 0.068 & -0.141 & 0.124\\
\hline
\end{tabular}

>>>>

\setstretch{1.5}
This document was last modified at 2025-11-21 21:08:43 and it was last
rendered at 2025-11-21 21:08:43.

@@

What is the legitimate scope of market inequality in the eyes of the
public? Since the early 1980s, many countries have experienced a
rollback of universal welfare programs and a shift toward the
privatisation and commodification of public goods and social services
(\citeproc{ref-gingrich_making_2011}{Gingrich, 2011};
\citeproc{ref-streeck_how_2016}{Streeck, 2016}). In Latin America, these
reforms extended market logic into domains of social reproduction long
governed by the state, shrinking public provision and expanding private
actors in core welfare sectors (\citeproc{ref-ferre_welfare_2023}{Ferre,
2023}). From a moral-economy perspective, this diffusion of market rules
has reconfigured the normative order: through policy feedback, welfare
institutions embed market-based criteria of ``fair'' allocation and
shape how citizens understand deservingness and the balance between
state and market (\citeproc{ref-koos_moral_2019}{Koos \& Sachweh, 2019};
\citeproc{ref-svallforsMoralEconomyClass2006a}{Svallfors, 2006}).
Against this backdrop, growing scholarship examines \emph{market justice
preferences}---the extent to which citizens regard market-based criteria
as a fair basis for access to essential services such as healthcare,
education and pensions
(\citeproc{ref-castillo_perceptions_2025}{Castillo, Laffert, et al.,
2025}; \citeproc{ref-lindh_public_2015}{Lindh, 2015};
\citeproc{ref-svallfors_political_2007}{Svallfors, 2007}). These
orientations matter because they help legitimise economic inequality by
portraying stratified outcomes as a fair result of individual
responsibility and limited state intervention.

This article examines public support for the commodification of pension
welfare---specifically, whether people consider it fair that access to
better pensions depends on income or contributions. Pension
privatisation lies at the core of Latin America's trajectory of
commodification (\citeproc{ref-huber_political_2000}{Huber \& Stephens,
2000}). As many countries expanded mandatory or voluntary
individual-capitalisation schemes---most radically in Chile---pensions
became organised around contribution-based entitlements and market
performance (\citeproc{ref-arenas_sistemas_2019}{Arenas, 2019};
\citeproc{ref-verbic_political_2019}{Verbič \& Spruk, 2019}). This shift
reshapes how pension institutions allocate old-age risk, tying benefit
levels to lifetime labour-market trajectories and increasing exposure to
poverty in retirement
(\citeproc{ref-madero-cabib_private_2019}{Madero-Cabib et al., 2019};
\citeproc{ref-oecd_pensions_2023}{OECD, 2023}). It also transforms the
justice principles citizens associate with old-age security: they must
judge whether higher benefits should follow market-derived criteria or
solidaristic principles
(\citeproc{ref-arrizabalomontoro_debate_2019}{Arrizabalo Montoro et al.,
2019}; \citeproc{ref-borzutzky_pension_2012}{Borzutzky, 2012}). Although
we know something about what people consider a ``just'' monetary pension
(\citeproc{ref-castillo_deserving_2019}{Castillo, Olivos, et al., 2019})
and about support for public versus private provision
(\citeproc{ref-busemeyer_welfare_2020}{Busemeyer \& Iversen, 2020};
\citeproc{ref-jaime-castillo_public_2013}{Jaime-Castillo, 2013}), we
still lack systematic evidence on how citizens assess the fairness of
market-dependent access to pensions.

Research shows that support for market-based access to welfare services
is strongly stratified: individuals with higher income, education or
class position are more likely to endorse market-based criteria than
disadvantaged groups (\citeproc{ref-busemeyer_skills_2014}{Busemeyer,
2014}; \citeproc{ref-immergut_it_2020}{Immergut \& Schneider, 2020};
\citeproc{ref-lindh_public_2015}{Lindh, 2015};
\citeproc{ref-svallfors_political_2007}{Svallfors, 2007}). Evidence for
pensions points in the same direction: higher income and greater
financial gains in funded schemes correlate with stronger support for
market-oriented pension policies
(\citeproc{ref-kerner_pension_2020}{Kerner, 2020}). Yet we know very
little about how movement within the class structure shapes preferences
for pension commodification. This gap is especially salient in Latin
America, where mobility has unfolded under high inequality and deeply
privatised welfare regimes
(\citeproc{ref-lopez-roldan_comparative_2021}{López-Roldán \& Fachelli,
2021}). Social origins and destinations anchor material interests and
justice principles
(\citeproc{ref-alesina_intergenerational_2018}{Alesina et al., 2018};
\citeproc{ref-gugushvili_subjective_2017}{Gugushvili, 2017};
\citeproc{ref-langsaether_explaining_2022}{Langsæther et al., 2022}),
while movement between these positions exposes individuals to new risks,
resources and normative environments that reshape fairness judgements
(\citeproc{ref-gugushvili_trends_2014}{Gugushvili, 2014};
\citeproc{ref-helgason_longterm_2023}{Helgason \& Rehm, 2023}). Social
mobility can thus foster market-justice orientations, helping explain
how inequalities in access to welfare services become normatively
sustained (\citeproc{ref-mau_inequality_2015}{Mau, 2015}).

Existing accounts identify several mechanisms through which
intergenerational mobility may shape distributive-justice preferences
but converge on two channels particularly relevant for market-based
welfare: changes in material stakes and shifts in normative frames
(\citeproc{ref-ares_changing_2020}{Ares, 2020};
\citeproc{ref-gugushvili_intergenerational_2016c}{Gugushvili, 2016a};
\citeproc{ref-helgason_longterm_2023}{Helgason \& Rehm, 2023};
\citeproc{ref-jaime-castillo_social_2019}{Jaime-Castillo \&
Marqués-Perales, 2019}). Recent research shows that meritocratic
beliefs---seeing that effort and ability are rewarded---are strongly
associated with support for market allocation
(\citeproc{ref-castillo_perceptions_2025}{Castillo, Laffert, et al.,
2025}). Perceived upward mobility reinforces meritocratic
interpretations of success, whereas downward mobility is more closely
linked to structural explanations of inequality without necessarily
eroding meritocratic convictions (\citeproc{ref-mijs_belief_2022}{Mijs
et al., 2022}). Thus, meritocracy and mobility operate on partly
separate tracks yet intersect in systematic ways. I treat meritocratic
beliefs as a post-treatment moderator---a diagnostic lens on potential
mechanisms. If mobility partly shapes preferences by strengthening
merit-based understandings of fairness and altering material stakes, its
effects should be most visible---and most pro-market---among individuals
who already interpret social outcomes through meritocratic lenses.

This article pursues two aims: first, to assess whether
intergenerational occupational mobility causally affects support for the
commodification of pensions; and second, to examine whether this effect
varies by meritocratic beliefs as a diagnostic of possible transmission
channels. I argue that social mobility operates through two main
pathways. Upward mobility increases earnings, expected returns in a
funded system and perceived economic security, making contribution-based
access more appealing. At the same time, exposure to higher-class
networks reinforces individualistic and effort-based understandings of
fairness. Downward mobility tends to heighten economic insecurity,
dependence on public protection and attachment to external attributions
of poverty, reducing support for market allocation. Meritocratic beliefs
are conceptualised not as mediators but as moderators. Among individuals
with stronger meritocratic beliefs, upward mobility should heighten
support for market-based access. In contrast, downward mobility should
weaken---or even reverse---its usual de-commodifying pattern, as
meritocratic interpretations of effort lead individuals to view market
outcomes as fair regardless of direction. The paper contributes to
research on market justice by identifying the causal effects of mobility
and by using meritocracy to probe whether these effects align with a
meritocratic channel.

The study focuses on Chile, an instructive case for examining
preferences for pension commodification. Despite sustained economic
growth, Chile combines high and persistent inequality with short-range
upward mobility and substantial barriers to higher-class positions
(\citeproc{ref-flores_top_2020}{Flores et al., 2020};
\citeproc{ref-salgado_uplifting_2025}{Salgado et al., 2025};
\citeproc{ref-torche_intergenerational_2014}{Torche, 2014}). It was the
first country worldwide to fully replace a public pay-as-you-go system
with a mandatory, privately administered defined-contribution scheme,
while the state has progressively expanded a segmented solidarity pillar
to compensate for insufficient benefits
(\citeproc{ref-boccardo_30_2020}{Boccardo, 2020};
\citeproc{ref-madariaga_three_2020}{Madariaga, 2020};
\citeproc{ref-solimano_rise_2021}{Solimano, 2021}). In parallel---and
despite waves of protest from 2006 to 2019 against the private pension
system (\citeproc{ref-somma_no_2021}{Somma et al., 2021})---Chilean
subjectivities have been increasingly shaped by neoliberal discourses
and market logics, influencing attitudes toward pension distribution and
inequality (\citeproc{ref-araujo_desafios_2012}{Araujo \& Martuccelli,
2012}; \citeproc{ref-galvez_con_2023}{Gálvez, 2023}). The combination of
profound inequality, market-driven old-age security and contested
legitimacy makes Chile a critical case for analysing how mobility and
meritocratic beliefs structure support for pension commodification.

To address these questions, the study draws on large-scale,
representative survey data for the urban Chilean population collected in
2016, 2018 and 2023. It adopts a causal framework of mobility effects
inspired by Breen \& Ermisch (\citeproc{ref-breen_effects_2024}{2024}).
The next section outlines the theoretical framework linking preferences
for market-based welfare, social mobility and expected heterogeneity by
meritocratic beliefs and time, and derives hypotheses. A subsequent
section presents the causal identification strategy, the data and the
analytical approach. The empirical sections report the findings, and the
article concludes by discussing what these results reveal about the
politics of market-based pension welfare and social mobility.

@@

\subsection{Preferences for market-based
welfare}\label{preferences-for-market-based-welfare}

@@

To explore whether mobility effects vary across key contextual
dimensions, I examine meritocratic perceptions as a moderator.
Meritocracy is captured with two items
(\citeproc{ref-young_rise_1958}{Young, 1958}), one referring to effort
(``In Chile, people are rewarded for their efforts'') and one to talent
(``In Chile, people are rewarded for their intelligence and skills''),
combined into a single index and dichotomized into low (≤ 3) and high (≥
4) meritocracy. Because meritocracy is measured after mobility, it is
used not as a causal effect modifier but as a post-treatment moderator
to probe potential mechanisms. I estimate interaction models between
mobility and meritocracy and report conditional treatment effects
(simple slopes) at low and high meritocratic levels.

@@

The specification is doubly robust: entropy-balancing weights are
combined with covariate adjustment so consistency holds if either the
weighting model or the outcome model is correctly specified, while also
improving efficiency
(\citeproc{ref-wooldridge_introductory_2009}{Wooldridge, 2009}). These
results are virtually identical to baseline models without covariates
(see \hyperref[anexo]{Supplementary Material} for complete models), with
no substantive changes in the mobility coefficients. Heterogeneous
effects by meritocratic perceptions are examined through interaction
models described in the previous section.

@@

\subsection{The role of meritocracy}\label{the-role-of-meritocracy}

To probe whether meritocratic beliefs channel the impact of mobility on
preferences for pension commodification, I estimate interaction models
between mobility and meritocratic beliefs and examine the conditional
treatment effects of mobility at low and high levels of meritocracy (see
\hyperref[anexo]{Supplementary Material} for complete models).

@@

Figure~\ref{fig-interact} displays the origin-specific ATT estimates
from the meritocracy interaction models as marginal effects of the
mobility treatment at low and high meritocratic beliefs. Across
trajectories, mobility effects for low-merit respondents are uniformly
small and imprecise, hovering around zero for both upward and downward
moves. Among high-merit respondents, the pattern aligns only partially
with the proposed mechanism: upward mobility from middle to high status
shows a clear and statistically significant pro-commodification effect
(\(\beta\) = 0.22, 95\% CI {[}0.05, 0.38{]}), whereas all other
trajectories---Low→Middle, Low→High, Middle→Low, High→Middle, and
High→Low---remain statistically indistinguishable from zero or display
negative estimates that do not systematically weaken under high merit.
Formal interaction tests also fail to detect strong heterogeneity by
meritocratic beliefs (see \hyperref[anexo]{Supplementary Material} for
interaction models). Taken together, the results provide only limited
support for the mechanism hypothesis: meritocracy appears to amplify the
pro-market consequences of upward mobility in one specific trajectory,
but there is no general or consistent evidence that meritocratic beliefs
constitute a stable channel through which intergenerational mobility
shapes preferences for pension commodification.

@@

\caption{\label{tbl-complete3}Interactions effects between
intergenerational occupational mobility and perceived meritocracy on
preferences for pension commodification, with covariates and wave fixed
effects.}

@@

\begin{center}
\scalebox{0.7}{
\begin{tabular}{l c c c c c c}
\hline
 & Low-Middle & Low-High & Middle-High & Middle-Low & High-Middle & High-Low \\
\hline
Intercept                                                    & $0.15$           & $-0.04$          & $0.30^{*}$       & $0.21$           & $0.08$           & $0.05$           \\
                                                             & $ [-0.36; 0.67]$ & $ [-0.40; 0.31]$ & $ [ 0.04; 0.56]$ & $ [-0.09; 0.51]$ & $ [-0.25; 0.40]$ & $ [-0.30; 0.41]$ \\
Mobility treatment                                           & $-0.02$          & $-0.03$          & $0.07$           & $-0.00$          & $-0.07$          & $-0.08$          \\
                                                             & $ [-0.09; 0.04]$ & $ [-0.10; 0.05]$ & $ [-0.01; 0.14]$ & $ [-0.07; 0.06]$ & $ [-0.15; 0.00]$ & $ [-0.15; 0.00]$ \\
High meritocracy perception (Ref.= Low)                      & $0.05$           & $0.06$           & $-0.01$          & $0.03$           & $0.17^{*}$       & $0.26^{*}$       \\
                                                             & $ [-0.06; 0.16]$ & $ [-0.07; 0.19]$ & $ [-0.09; 0.08]$ & $ [-0.07; 0.13]$ & $ [ 0.00; 0.33]$ & $ [ 0.04; 0.48]$ \\
Mobility treatment x High meritocracy perception (Ref.= Low) & $0.04$           & $0.05$           & $0.15$           & $0.04$           & $-0.03$          & $-0.14$          \\
                                                             & $ [-0.12; 0.20]$ & $ [-0.14; 0.24]$ & $ [-0.03; 0.33]$ & $ [-0.11; 0.19]$ & $ [-0.23; 0.17]$ & $ [-0.38; 0.10]$ \\
\hline
Controls                                                     & Yes              & Yes              & Yes              & Yes              & Yes              & Yes              \\
R$^2$                                                        & $0.04$           & $0.05$           & $0.08$           & $0.04$           & $0.07$           & $0.10$           \\
Adj. R$^2$                                                   & $0.03$           & $0.04$           & $0.07$           & $0.02$           & $0.06$           & $0.08$           \\
Num. obs.                                                    & $878$            & $745$            & $716$            & $773$            & $912$            & $861$            \\
RMSE                                                         & $0.40$           & $0.37$           & $0.39$           & $0.39$           & $0.44$           & $0.41$           \\
N Clusters                                                   & $497$            & $449$            & $443$            & $470$            & $536$            & $525$            \\
\hline
\multicolumn{7}{l}{\scriptsize{Note: Cells contain regression coefficients with confidence intervals in parentheses. $^*$ Null hypothesis value outside the confidence interval..}}
\end{tabular}
}
\label{table:coefficients}
\end{center}

@@

\caption{\label{tbl-complete4}Marginal effects of mobility by
meritocratic perception (Low vs High Meritocracy).}

@@
